Think deep for reef resilience

### 2025-07-11 paper/paper_draft.qmd
Add references

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes
[@grimaldi2022]. Current modelling projects severe declines in coral cover under projections
of Degree Heating Weeks \[**Bozec 2025**\]. These coral cover declines will be compounded by
ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the
2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with
some level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

@@

While these environmental and biological factors determine whether individual corals survive
thermal stress events, the long-term viability of coral reefs depends not only on coral
survival but also on the maintenance of reef structure through positive carbonate budgets.
Scleractinian corals are primary habitat structures on coral reefs, and the accumulation and
erosion of coral reef carbonate frameworks is controlled by a range of processes including
coral accumulation \[**Perry**\], bioerosion , physical damage generating coral rubble
\[**Kenyon**\], and calcification and CaCO₃ dissolution rates \[**Eyre 2014, 2018**\]. The
balance of these processes and occurrence of net carbonate framework production (referred to
as a positive carbonate budget) are influenced by factors such as coral cover and complexity
\[**Perry, Desbiens**\], rates of physical environmental processes, and ocean acidity
\[**Perry, Eyre**\]. Total live coral cover on a reef highly influences the maintenance of
net positive carbonate budgets, with estimates ranging from 10 to 20% \[**Perry**\] of a
reef's total area required to maintain positive budgets. For mid and outer shelf reefs on
the Great Barrier Reef, this threshold is approximately 17% live coral cover
\[**Desbiens**\].

>>>>

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes
[@grimaldi2022]. Current modelling projects severe declines in coral cover under projections
of Degree Heating Weeks [@bozec]. These coral cover declines will be compounded by ocean
acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the
2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with
some level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

@@

While these environmental and biological factors determine whether individual corals survive
thermal stress events, the long-term viability of coral reefs depends not only on coral
survival but also on the maintenance of reef structure through positive carbonate budgets.
Scleractinian corals are primary habitat structures on coral reefs, and the accumulation and
erosion of coral reef carbonate frameworks is controlled by a range of processes including
coral accumulation [@perry2015], bioerosion , physical damage generating coral rubble
[@kenyon], and calcification and CaCO₃ dissolution rates [@eyre2014; @eyre2018]. The balance
of these processes and occurrence of net carbonate framework production (referred to as a
positive carbonate budget) are influenced by factors such as coral cover and complexity
[@perry2012; @desbiens], rates of physical environmental processes, and ocean acidity
[@perry2015; @eyre2018]. Total live coral cover on a reef highly influences the maintenance
of net positive carbonate budgets, with estimates ranging from 10 to 20% [@perry2015] of a
reef's total area required to maintain positive budgets. For mid and outer shelf reefs on
the Great Barrier Reef, this threshold is reportedly 17% live coral cover [@desbiens].
